\myTitle

### 2025-03-25 tex/responses-to-referees/responses-to-reviewers-1.Rmd

* Authors provide considerable detail on sample location in the methods but only provide the sample number in the introduction. I believe it would be useful to have summary sentences of Table 1 that include number sites for spring, fall, etc run, and the number of samples within each ecotype, total samples within the methods, etc.
    - _Thank you for this suggestion.  We added the following to the end of the first paragraph in the Population Sampling portion of the
Methods section:_
"Of the 1636 total fish samples from 17 different populations, 727 fish from 10 collections were of the fall run ecotype,
498 fish from 5 collections were spring run, and 111 and 300 fish were of the winter run and late-fall run ecotypes, 
respectively,  each represented by a single collection."

* Is there any LOD threshold for your GWAS that allowed identifying the chr34 peak? 
    - _We updated our treatment of this topic to note that we relied upon the commonly used
    GWAS $p$-value threshold of $5 \times 10^{-8}$.  We also added an additional
    supplemental figure that shows all the candidates chosen for followup within and
    near that peak._

* In addition, L239-240 implies you were you only interested in right side of the peak. I would assume you were interested in up and downstream of the peak unless I missed something. In addition, please clarify why 2.2 Mb selected. Is this an LD block? Does this mean 2.2 Mb in each direction or total?
    - _It turns out that the main peak of association is at the
    beginning of chromosome 34, so there is no flanking region to the left of the peak. To
    clarify why those candidate SNPs in the flanking region were chosen we now state:_ 
    "The final condition was implemented in order to gather several SNPs with large allele frequency differences
that were adjacent to, but not directly within, the main peak of association, so as to possibly learn about
recombination in the region. "

@@

* The writing of the manuscript is largely well written but makes high use of words such as “we” or “us”, which although stylistic, makes the manuscript longer than needed in some cases such as “we then dropped” or “this allows us”. In many cases these can be removed (examples in minor comments).
    - _We addressed all the examples in the minor comments and also removed additional instances of the
    use of such language where it was practical._

* There is also use of informal language e.g. “amplified so well” or unnecessary words such as “then” that could be removed.
    - _Following this suggestion, we removed many occurrences of "then" and "as well as"_

* I commend the use of the distribution of log-likelihood ratio for relationship to show the power of haplotypes vs SNPs and would argue this could be elevated to a full figure, potentially in a condensed form, maybe less panels somehow?
    - _We appreciate the referee's feedback on this.  In response, we have created a
    new figure in the paper itself that examines the distribution of log-likelihood
    ratios only in the Feather River Hatchery.  This figure distinguishes between
    different log-likelihood ratios (PO/U, FS/U, and HS/U) and nicely shows, in a graphical way, the
    advantages of using microhaplotypes for relationship inference. We have also left
    the original figure in the supplement for any readers curious about the effect
    in different collections._

@@

* If possible, I think it would also be good to show how much of an improvement the discrimination power improved from the old marker data. In theory, this data has already been generated for the full population if you strip out the novel SNPs of the current study.
    - _After long and careful consideration of this, we determined that it would not
    be possible to make an accurate comparison between the old and the new marker
    data in this regard.  Unfortunately, only 78 of the original 94 TaqMan assays
    were able to be converted into amplicon-sequenced markers, so we are not able
    to reproduce the old marker set with the same reference samples._

* It may also be worth discussing other studies on other organisms that either support or do not support the utilization and increased power achieved through haplotypes over SNPs.
    - _We have included four additional citations in the first paragraph of the discussion
    to place this work in the context of those other investigations._

@@

- L296 & L317: I’m not sure of the difference between the microhaplotype markers and SNPlicon targets. I presumed they are the same thing after it was stated that microhaplotypes would “hereafter referred to as SNPlicons” but this term is rarely used again, please decide whether you want to keep this term and avoid using both names or clarify if these are different.
    - _We updated the introduction of the term SNPlicon to specifically state that it refers to the microhaplotypes
    that were obatined from sequencing around the sites of the Taq-Man assays that were previously used.
    We feel it is quite important to retain information about the provenance of these markers as it could
    facilitate merging amplicon-sequenced data with previously collected Taq-Man data, so we have retained
    the terminology._

- L316: Remove the double bracket/parentheses
    - _Done_

- L319: Package references only need to used on the first mention of the package
    - _Eliminated._

- L386-390: I had a hard time following this sentence, please clarify.
    - _This was just a technical statement about the way that CKMRsim works because
    of the fact that physical linkage in the absence of LD has no effect on the likelihoods
    for parent-offspring pairs and unrelated pairs.  However since simulation with
    physical linkage in PO and U relationships yields identical results to those without
    physical linkage, we just simplify the manuscript by not mentioning that technical
    detail._

@@

## Reviewer 2

This study develops a microhaplotype panel (based on the GT-seq method) to create a training set to identify different populations of genetic stock and relationships between individuals in America Salmon. My comments relating to this manuscript are outlined below:

### Major Comments

* Lines 399-444: My biggest compliant is the paper states that it creates a “baseline” for GSI and PBT analysis but it is not always clear what this final baseline is and only at the end of the introduction does it state what this baseline contains (in terms of number of animals and number of loci). I feel this needs to also be stated in the abstract. 
    - _As per reviewer \#1's suggestion, the methods now includes textual comment on the
    number of samples from each collection. Additionally, we have added several sentences
    in the third paragraph of the introduction to define the terms "reference data set" and
    "baseline," and we have also added a summary in the abstract
    on the total number of amplicons, individuals and populations in the reference data set._

@@

* In addition, in the results section it is not easy for work out how many markers made into the final reference baseline, and I think this information should be included in a table that also shows how markers from each source of information. I’m assuming that this is what is being used in the subsequent analysis.
    - _Excellent suggestion.  This information now appears in the new Table 2 which shows the total number of
    amplicons and the total number of variant sites within each amplicon for each of several
    groups of markers that correspond to the different sources of information._

* Table 3: The values in second column (the average number of fish with observed genotypes) seem odd to me. I would have thought these values would be less than seventh column in Table 1 (column N). For example, CHLF has 379.6 in Table 3 > 261 in Table 1. Can the authors explain this? Furthermore, the proportion of individuals with observed genotypes is a more useful metric than the average number of non-missing genotypes, since there the number of fish between sampling sites is very different.
    - _Thank you very much for noting this.  We have now corrected this problem. The table was
    created from an old version of our metadata file that was missing entries from 42 fish.  These
    fish are now properly included in the metadata file and the quantities in Table 1 have been
    regenerated. Also, we agree with the referee that the average number of non-missing genotypes
    is a less useful metric than one quantified as a proportion.  Therefore we have replaced
    the average number of non-missing genotypes with the total proportion of missing data in Table 1._

@@

- Line 48: “… West Coast” should be “West Coast of the USA”?
    - _Fixed_

- Line 66: The term “reference baselines” is used through the manuscript but it is quite vague what this means. Reading the manuscript, it would seem to me that the reference baseline has two parts (1) “a multihaplotype marker/SNP set” and (2) “a reference dataset for training the GSI and PBT methods”. I think the authors define the term “reference baselines” here in the introduction.
    - _Thank you.  The paper really needed this.  We now have a paragraph in the 
    introduction that defines these terms_

- Introduction – The paper seems to rely heavily on the work by Thompson et al (2020) in the methods, but this is not mentioned in the introduction. The authors should make a comment towards of the introduction that the work depends on some of the data and results from this paper.
    - _Added a clause in the last paragraph of the introduction about this._

- Line 181: "a de novo assembly was performed". How was this performed? More details are needed.
    - _We clarified that we used ustacks from the Stacks program to assemble the reads into
    RAD loci._

@@

- Line 516: “… half of those (10) were the incorrect assignments”. I found this a little confusing. Are you saying that 10 of the discarded assignments were from the 24 incorrect assignments reported in line 501? This sentence could possibly do with a rephasing.
    - _Edited for clarity_

- Caption of Table 3: “N is the number of fish …” should be “N is the average number of fish …”
    - _Fixed_

- Figure 1 and 2: The colours for “spring run” and “Late-fall run” are very difficult to separate for a colour-blind person. Since the groups are not in a specific order on the graph, it is very hard work to out what is what. I’d recommend the authors use a selection of colours that will work for colour-blind people.

    - _We greatly appreciate this feedback and apologize for the difficulty in
    discerning those two colors. We have introduced several changes to the figures
    that we hope will make the distinction between those colors much easier.
    In order to keep the color scheme consistent, as much as possible,
    with previous publications we have employed different shapes and striping to
    make the different run types more distinct for colour-blind people. Specifically,
    Figure 1 (the map) now uses different colors and shapes for the different run
    timing groups.  Figure 2 uses striping across the Late-fall run group sectors
    to clearly distinguish that color from the spring-run color._

### 2025-03-25 tex/abstract.tex
Edits ID'd looking at the latexdiff

Genetic methods have become an essential component of ecological investigation
and conservation planning for fish and wildlife. Among these methods is the use
of genetic marker data to identify individuals to
populations, or stocks, of origin. More recently, methods that involve genetic
pedigree reconstruction to identify relationships between individuals within populations
have also become common. We present, here, a novel set of
multi-allelic microhaplotype genetic markers for Chinook salmon which provide unprecedented
resolution for population discrimination and relationship identification from a rapidly and
economically assayed panel of markers. We show how this set of microhaplotypes
provides definitive power to identify all known lineages of Chinook salmon in
California. The inclusion of genetic loci that have known associations with phenotype
and that were identified as outliers in examination of whole genome sequence data,
allows resolution of stocks that are not highly genetically differentiated but
are phenotypically distinct and managed as such. 
This same set of multiallelic genetic markers have ample variation to
accurately identify parent-offspring and full-sibling
pairs in all California populations, including the genetically depauperate winter-run
lineage.
Validation of this marker panel in coastal salmon populations not previously studied with
modern genetic methods, also reveals novel biological insights, including
the presence of a single copy of a haplotype for a phenotype that has not
been documented in that part of the species range, and a clear signal of mixed
ancestry for a salmon population that is on the geographic margins of the 
primary evolutionary lineages present in California.

>>>>

Genetic methods have become an essential component of ecological investigation
and conservation planning for fish and wildlife. Among these methods is the use
of genetic marker data to identify individuals to
populations, or stocks, of origin. More recently, methods that involve genetic
pedigree reconstruction to identify relationships between individuals within populations
have also become common. We present, here, a novel set of
multi-allelic microhaplotype genetic markers for Chinook salmon which provide unprecedented
resolution for population discrimination and relationship identification from a rapidly and
economically assayed panel of markers. We show how this set of genetic markers assayed
by sequencing of 204 amplicons, in tandem with a reference data set of 1636 individual
samples from 17 populations, 
provides definitive power to identify all known lineages of Chinook salmon in
California. The inclusion of genetic loci that have known associations with phenotype
and that were identified as outliers in examination of whole genome sequence data,
allows resolution of stocks that are not highly genetically differentiated but
are phenotypically distinct and managed as such. 
This same set of multiallelic genetic markers have ample variation to
accurately identify parent-offspring and full-sibling
pairs in all California populations, including the genetically depauperate winter-run
lineage.
Validation of this marker panel in coastal salmon populations not previously studied with
modern genetic methods, also reveals novel biological insights, including
the presence of a single copy of a haplotype for a phenotype that has not
been documented in that part of the species range, and a clear signal of mixed
ancestry for a salmon population that is on the geographic margins of the 
primary evolutionary lineages present in California.

### 2025-03-25 tex/body-text-of-paper.tex
Edits ID'd looking at the latexdiff

We present a new reference baseline that includes 1,636 Chinook salmon individuals from 17 collections that were genotyped at 204 loci distributed throughout the genome.
This baseline is highly effective for GSI within California and provides ample power for PBT, thus enabling a highly effective and efficient integrated GSI/PBT monitoring and evaluation program 
\citep{garza2007large,beacham2021parentage}.
We describe how we identified and compiled the markers, including several new markers
identified through re-analysis of the whole genome sequencing data of \citet{thompson2020complex},
that are particularly effective at distinguishing between closely related groups of fish,
such as late-fall and fall-run Chinook salmon in the CCV.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.

@@

Candidate loci with multiple single-nucleotide polymorphisms (SNPs) among multiple
populations were identified using reduced-representation genome sequencing of 2--4 Chinook salmon, each, from a number of
populations ranging from California to British
Columbia and the interior Columbia River (Table~\ref{tab:ascert-panel}).
Genomic DNA was extracted using Qiagen DNeasy 96 Blood and Tissue Kits on a BioRobot 3000
(Qiagen, Inc). Double-digest restriction-site associated DNA sequencing (ddRAD) with different size selections (400 bp and 500 bp
fragments) was used to produce a broad array of locus targets. Library preparation and sequencing methods followed those of
\citet{peterson2012double} with the modifications described in \citet{baetscher2018microhaplotypes}.
Sequencing was performed on a MiSeq (Illumina, Inc) using $2 \times 300$-cycle paired-end sequencing.
Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, the ustacks module of Stacks were used to assemble reads {\em de novo} into RAD loci.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.

The 8,725 discovered RAD loci were filtered down to a set of markers suitable for conversion
to amplicon sequencing assays. Gene regions were selected from the RAD loci that (1) contained multiple SNPs within a 100 bp window,
and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is
diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.
These 229 gene regions were then tested for PCR-amplification and consistency in a separate sample
of 288 Chinook salmon: 96 SRW fish and 48 each from the Entiat River (a Columbia River
tributary), TRH, FRHS, and CRLF\@.  Eight GT-seq amplifications were made and these were sequenced
on eight runs of a MiSeq sequencer
using a $2\times 75$-cycle kit to identify problematic loci and to optimize primer dilutions.
The sequences from each run at each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignments were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

A number of SNP markers from an earlier discovery effort \citep{clemento2011discovery} had already
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp., Carlsbad, CA) probes or SNPtype assays (Standard BioTools Inc., South San Francisco, CA).
These markers were typed using amplicon sequencing, following the methods in
\citet{campbell2015genotyping} with the addition of the Illumina small RNA sequencing primer and
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus
Specific Primers respectively.  Loci were removed that were frequently inconsistent with SNPtype assays,
had poor amplification, or had excessive amplification and were subsequently over-represented in the sequencing library.  FreeBayes \citep{garrison2012haplotype} was used to identify all SNPs (additional to the TaqMan-assay targets) along the fragment, which facilitated scoring fragment microhaplotypes when multiple SNPs were present along the read \citep{baetscher2018microhaplotypes}. The resulting genetic markers are hereafter referred to as `SNPlicons' to record their origin as TaqMan assays in a previous
baseline.

@@

The novel microhaplotype markers and the SNPlicon targets were
developed prior to the advent of a well-assembled genome for Chinook salmon.
As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both of these
groups of markers we identified their locations within the genome by mapping the
consensus sequences (roughly 100--300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0, GenBank accession GCA\_002872995.1)
and version 2  (Otsh\_v2.0, GenBank accession GCA\_018296145.1) genome assemblies,
using both bwa mem \citep{bwa-mem2009} and the BLAST-like Alignment tool, BLAT \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers to
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

Finally, we investigated the matrix of assignments and misassignments
broken down according to the genotype of the fish at the markers within
the RoSA.  RoSA genotype \citep[EE, EL, or LL, see~][]{thompson2020complex}
was assessed from the concordance of the genotypes
at all 8 markers described in \citep{thompson2020complex}, requiring that no genotypes be missing
within an individual from any of those 8 markers, and discarding data from 12 individuals
in which the E or L haplotypes had clearly recombined.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize  Summary of amplicons in the reference data set.  Source denotes
which discovery effort yielded the amplicon. N (amplicon) is the number of amplicons
and N (variant) is the total number of variant sites assayed within the amplified sequence of all the
amplicons.  Source abbreviations are as follows: mhap = novel microhaplotype discovery, scon = conversion of existing SNPtype
assays (the `SNPlicons'), rosa = adult-migration-timing associated markers, wrap = winter-run associated
polymorphisms, lfar = late-fall-run associated polymorphisms, sixx and vgll = amplicons
targeting the SIX6 and VGLL3 genes, respectively, sexy = sexID marker, coho =
distinguishes coho from Chinook salmon. The $^*$ denotes markers routinely
excluded from likelihood calculations for GSI.}
\label{tab:loci-summ}
{
\begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}} lrr}
\hline\hline
Type&N (amplicon)&N (variant)\tabularnewline
\hline
\input{inputs/amplicon-summary.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule width0.48\textwidth
\vspace*{0.3ex}\hrule width0.48\textwidth
\end{table}
%%%%%%%%%%%%%%%

@@

Of 229 tested candidates from our novel microhaplotype discovery process
125 markers were retained \citep{thompson2020complex}
for use in stock identification within
the Klamath Basin.  However, further evaluation of those markers in other
populations led the removal of
twelve loci with apparent null alleles in the winter-run population, two that were monomorphic
in the winter-run, despite having
multiple alleles in other populations, and an additional 5 loci that were
difficult to score in routine application of the baseline. This left 106
novel microhaplotype amplicons in the current panel.

@@

Analysis of the mapping of amplicons for Chr34:828,768
showed that many ($\approx$89\%) of the reads were off-target (aligning to other chromosomes,
etc.), making it costly (in terms of sequencing effort) to include the marker in the baseline panel.
Consequently it was dropped from the panel.  The two remaining loci, which are at coordinates
Chr34:865,057, and Chr34:1,063,084 in the Otsh\_v1.0 assembly are at coordinates
Chr34:954,054 and  Chr34:1,151,868 in the Otsh\_v2.0 assembly, and
the frequency of the different alleles at the two remaining loci
within different reporting units in the baseline are shown in Table~\ref{tab:lfar-freqs}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Allele frequencies across reporting units of the two late-fall associated
markers.  Frequencies are given for the alleles most common amongst late-fall Chinook: nucleotide
{\tt G} at locus Chr34:954,054 and {\tt A} at Chr34:1,151,868.  $N$ is the total number of gene
copies available for estimating the allele frequency. Genome coordinates of SNPs are from Otsh\_v2.0.}
\label{tab:lfar-freqs}
{\footnotesize
\begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}} lrrrr}
\hline\hline
& \multicolumn{2}{c}{\underline{Chr34:954,054}} & \multicolumn{2}{c}{\underline{Chr34:1,151,868}} \\
Reporting Unit & freq. & $N$ & freq. & $N$ \\ \hline
\input{inputs/lfar-freqs.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule width0.48\textwidth
\vspace*{0.3ex}\hrule width0.48\textwidth
\end{table}
%%%%%%%%%%%%%%%

@@

The population-genetic summary statistics are presented in Table~\ref{tab:pg-summ}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Simple population genetic summaries. All quantities are means
over loci.   $M$ is the total fraction of missing genotypes; $\bar{N}_\mathrm{A,ss}$ is the
average number of alleles and $\bar{P}_\mathrm{poly,ss}$ the fraction of polymorphic loci,
after sub-sampling to the smallest sample size per locus; $\bar{H}_\mathrm{exp}$ and
$\bar{H}_\mathrm{obs}$ are expected and observed heterozygosity, respectively. Population
codes are as given in Table~\ref{tab:samples}}
\label{tab:pg-summ}
{\footnotesize
\begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}} lrrrrr}
\hline\hline
\input{inputs/popgen-summary.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule width0.48\textwidth
\vspace*{0.3ex}\hrule width0.48\textwidth
\end{table}
%%%%%%%%%%%%%%%
The total fraction of missing data by collection varied from a low of
0.008  in CRHS to a high of 0.338 in SRW, the latter due to a tranche of
samples from degraded carcasses. The average number of alleles
per locus, standardized to the smallest sample size, ranged from 2.32 (SRW) to
2.64 (RRF), with  SRW also having the smallest value of the
standardized fraction of polymorphic markers at 0.88 and RRF sharing the highest value
(0.96) with FHRF and BCkF.
SRW also had the lowest values for expected and observed heterozygosity:
0.338 and 0.329, respectively. While the highest for expected heterozygosity
was 0.401 (SmRF) and for observed heterozygosity, 0.399 (CRHS).

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that provides sufficient power for highly accurate identification 
of parent and offspring pairs and full siblings, and also allows near-perfect
identification of individuals to population or genetic group of origin in California. 
This set of genetic markers includes multiallelic gene regions with high variability for
relationship inference, gene regions identified in whole-genome sequence data
for increased power for identification of specific populations, and markers that have
long been in use for both GSI and PBT in California \citep{clemento2014evaluation}, by converting them
into so-called 'SNPlicons,' which are treated as multiallelic haplotypes in the presence
of additional variation surrounding the originally assayed SNP. Our work adds to a growing
chorus of recent studies finding that microhaplotype markers can provide additional
power over single SNPs for both GSI \citep{mckinney2017managing,mckinney2022new,hargrove2024efficient} and
relationship inference \citep{baetscher2018microhaplotypes,bootsma2020gt,delomas2024microhaplotypes}.
This set of markers lays the foundation for a comprehensive 
genetic monitoring and evaluation effort that facilitates multiple types of inference 
and is flexible and extensible.

>>>>

We present a new reference baseline that includes 1,636 Chinook salmon individuals from 17 collections that were genotyped at 204 loci distributed throughout the genome.
This baseline is highly effective for GSI within California and provides ample power for PBT inference of other close relatives, thus enabling a highly effective and efficient integrated GSI/PBT monitoring and evaluation program 
\citep{garza2007large,beacham2021parentage}.
We describe how we developed and compiled the markers, including several new markers
identified through re-analysis of the whole genome sequencing data of \citet{thompson2020complex},
that are particularly effective at distinguishing between closely related groups of fish,
such as late-fall and fall-run Chinook salmon in the CCV.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.

@@

Candidate loci with multiple single-nucleotide polymorphisms (SNPs) among multiple
populations were identified using reduced-representation genome sequencing of 2--4 Chinook salmon, each, from a number of
populations ranging from California to British
Columbia and the interior Columbia River (Table~\ref{tab:ascert-panel}).
Genomic DNA was extracted using Qiagen DNeasy 96 Blood and Tissue Kits on a BioRobot 3000
(Qiagen, Inc). Double-digest restriction-site associated DNA sequencing (ddRAD) with different size selections (400 bp and 500 bp
fragments) was used to produce a broad array of locus targets. Library preparation and sequencing methods followed those of
\citet{peterson2012double} with the modifications described in \citet{baetscher2018microhaplotypes}.
Sequencing was performed on a MiSeq (Illumina, Inc) using $2 \times 300$-cycle paired-end sequencing.
Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, the ustacks module of Stacks was used to assemble reads {\em de novo} into RAD loci.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.

The 8,725 discovered RAD loci were filtered down to a set of markers suitable for conversion
to amplicon sequencing assays. Gene regions were selected from the RAD loci that (1) contained multiple SNPs within a 100 bp window,
and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is
diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.
These 229 gene regions were tested for PCR-amplification and consistency in a separate sample
of 288 Chinook salmon: 96 SRW fish and 48 each from the Entiat River (a Columbia River
tributary), TRH, FRHS, and CRLF\@.  Eight GT-seq amplifications were made and these were sequenced
on eight runs of a MiSeq sequencer
using a $2\times 75$-cycle kit to identify problematic loci and to optimize primer dilutions.
The sequences from each run at each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler (bwa) aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignments were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

A number of SNP markers from an earlier discovery effort \citep{clemento2011discovery} had already
been validated as useful for GSI
and PBT in California \citep{clemento2014evaluation}. These markers were originally assayed
using TaqMan (Life Technologies Corp., Carlsbad, CA) probes or SNPtype assays (Standard BioTools Inc., South San Francisco, CA).
These markers were typed using amplicon sequencing, following the methods in
\citet{campbell2015genotyping} with the addition of the Illumina small RNA sequencing primer and
read 2 sequencing primer to the SNPtype Specific Target Amplification Primer and Locus
Specific Primers respectively.  Loci were removed that were frequently inconsistent with SNPtype assays,
had poor amplification, or had excessive amplification and were subsequently over-represented in the sequencing library.  FreeBayes was used to identify all SNPs (additional to the TaqMan-assay targets) along the fragment, which facilitated scoring fragment microhaplotypes when multiple SNPs were present along the read \citep{baetscher2018microhaplotypes}. The resulting genetic markers are hereafter referred to as `SNPlicons' to record their origin as TaqMan assays in a previous
baseline.

@@

The novel microhaplotype markers and the SNPlicon targets were
developed prior to the advent of a well-assembled genome for Chinook salmon.
As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both of these
groups of markers we identified their locations within the genome by mapping the
consensus sequences (roughly 100--300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0, GenBank accession GCA\_002872995.1)
and version 2  (Otsh\_v2.0, GenBank accession GCA\_018296145.1) genome assemblies,
using both bwa mem and the BLAST-like Alignment tool, BLAT \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers to
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

Finally, we investigated the matrix of assignments and misassignments
broken down according to the genotype of the fish at the markers within
the RoSA.  RoSA genotype \citep[EE, EL, or LL, see][]{thompson2020complex}
was assessed from the concordance of the genotypes
at all 8 markers described in \citet{thompson2020complex}, requiring that no genotypes be missing
within an individual from any of those 8 markers, and discarding data from 12 individuals
in which the E or L haplotypes had clearly recombined.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize  Summary of amplicons in the reference data set.  Source denotes
which discovery effort yielded the amplicon. N (amplicon) is the number of amplicons
and N (variant) is the total number of variant sites assayed within the amplified sequence of all the
amplicons.  Source abbreviations are as follows: mhap = novel microhaplotype discovery, scon = conversion of existing SNPtype
assays (the `SNPlicons'), rosa = adult-migration-timing associated markers, wrap = winter-run associated
polymorphisms, lfar = late-fall-run associated polymorphisms, sixx and vgll = amplicons
targeting the SIX6 and VGLL3 genes, respectively, sexy = sexID marker, coho =
distinguishes coho from Chinook salmon. The $^*$ denotes markers routinely
excluded from likelihood calculations for GSI.}
\label{tab:loci-summ}
\begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}} lrr}
\hline\hline
Type&N (amplicon)&N (variant)\tabularnewline
\hline
\input{inputs/amplicon-summary.tex}
\end{tabular*}
\par\vspace*{-3.2ex}\noindent\rule{0.48\textwidth}{0.5pt}
\par\vspace*{-2.4ex}\noindent\rule{0.48\textwidth}{0.5pt}
\end{table}
%%%%%%%%%%%%%%%

@@

Of 229 tested candidates from our novel microhaplotype discovery process
125 markers were retained \citep{thompson2020complex}
for use in stock identification within
the Klamath Basin.  However, further evaluation of those markers in other
populations led to the removal of
twelve loci with apparent null alleles in the winter-run population, two that were monomorphic
in the winter-run, despite having
multiple alleles in other populations, and an additional 5 loci that were
difficult to score in routine application of the baseline. This left 106
novel microhaplotype amplicons in the current panel.

@@

Analysis of the mapping of amplicons for Chr34:828,768
showed that many ($\approx$89\%) of the reads were off-target (aligning to other chromosomes,
etc.), making it costly (in terms of sequencing effort) to include the marker in the baseline panel.
Consequently it was dropped from the panel.  The two remaining loci, which are at coordinates
Chr34:865,057, and Chr34:1,063,084 in the Otsh\_v1.0 assembly are at coordinates
Chr34:954,054 and  Chr34:1,151,868 in the Otsh\_v2.0 assembly, and
the frequency of the different alleles at the two remaining loci
within different reporting units in the baseline are shown in Table~\ref{tab:lfar-freqs}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Allele frequencies across reporting units of the two late-fall associated
markers.  Frequencies are given for the alleles most common amongst late-fall Chinook: nucleotide
{\tt G} at locus Chr34:954,054 and {\tt A} at Chr34:1,151,868.  $N$ is the total number of gene
copies available for estimating the allele frequency. Genome coordinates of SNPs are from Otsh\_v2.0.}
\label{tab:lfar-freqs}
{\footnotesize
\begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}} lrrrr}
\hline\hline
& \multicolumn{2}{c}{\underline{Chr34:954,054}} & \multicolumn{2}{c}{\underline{Chr34:1,151,868}} \\
Reporting Unit & freq. & $N$ & freq. & $N$ \\ \hline
\input{inputs/lfar-freqs.tex}
\end{tabular*}
}
\par\vspace*{-2.2ex}\noindent\rule{0.48\textwidth}{0.5pt}
\par\vspace*{-2.4ex}\noindent\rule{0.48\textwidth}{0.5pt}
\end{table}
%%%%%%%%%%%%%%%

@@

The population-genetic summary statistics are presented in Table~\ref{tab:pg-summ}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Simple population genetic summaries. All quantities are means
over loci.   $M$ is the total fraction of missing genotypes; $\bar{N}_\mathrm{A,ss}$ is the
average number of alleles and $\bar{P}_\mathrm{poly,ss}$ the fraction of polymorphic loci,
after sub-sampling to the smallest sample size per locus; $\bar{H}_\mathrm{exp}$ and
$\bar{H}_\mathrm{obs}$ are expected and observed heterozygosity, respectively. Population
codes are as given in Table~\ref{tab:samples}}
\label{tab:pg-summ}
{\footnotesize
\begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}} lrrrrr}
\hline\hline
\input{inputs/popgen-summary.tex}
\end{tabular*}
}
\par\vspace*{-2.2ex}\noindent\rule{0.48\textwidth}{0.5pt}
\par\vspace*{-2.4ex}\noindent\rule{0.48\textwidth}{0.5pt}
\end{table}
%%%%%%%%%%%%%%%
The total fraction of missing data by collection varied from a low of
0.008  in CRHS to a high of 0.338 in SRW, the latter due to a tranche of
samples from degraded carcasses. The average number of alleles
per locus, standardized to the smallest sample size, ranged from 2.32 (SRW) to
2.64 (RRF), with  SRW also having the smallest value of the
standardized fraction of polymorphic markers at 0.88 and RRF sharing the highest value
(0.96) with FHRF and BCkF.
SRW also had the lowest values for expected and observed heterozygosity:
0.338 and 0.329, respectively. While the highest for expected heterozygosity
was 0.401 (SmRF) and for observed heterozygosity, 0.399 (CRHS).

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that provides sufficient power for highly accurate identification 
of parent and offspring pairs and full siblings, and also allows near-perfect
identification of individuals to population or genetic group of origin in California. 
This set of genetic markers includes multiallelic gene regions with high variability for
relationship inference, gene regions identified in whole-genome sequence data
for increased power for identification of specific populations, and markers that have
long been in use for both GSI and PBT in California \citep{clemento2014evaluation}, by converting them
into so-called 'SNPlicons,' which are treated as multiallelic haplotypes in the presence
of additional variation surrounding the originally assayed SNP\@. Our work adds to a growing
chorus of recent studies finding that microhaplotype markers can provide additional
power over single SNPs for both GSI \citep{mckinney2017managing,mckinney2022new,hargrove2024efficient} and
relationship inference \citep{baetscher2018microhaplotypes,bootsma2020gt,delomas2024microhaplotypes}.
This set of markers lays the foundation for a comprehensive 
genetic monitoring and evaluation effort that facilitates multiple types of inference 
and is flexible and extensible.
